---
title: "Storm Data"
author: "patmau"
date: "6/1/2020"
output: html_document
---

# Storm Data Analysis

## Synopsis

bla bla bla

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## System info

```{r systemInfo}
sessionInfo()
```

## Download and process data

Download the file if necessary and load it into R:

```{r download}
dataFile <- "data/StormData.csv.bz2"
if (!file.exists(dataFile)) {
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(url, dataFile, method = "curl")   
}

data <- read.csv(dataFile, header = TRUE, stringsAsFactors = FALSE)
```

### Event types

The [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) lists 48 permitted event types.
In the data file, however, we find no less than `r length(levels(factor(data$EVTYPE)))` event types.
The top five most occurring event types:
```{r eventTypes}
head(summary(factor(data$EVTYPE)), 5)
```
Clearly, "TSTM WIND" and "THUNDERSTORM WIND" refer to the same type of event.
We perform a coarse clean-up of event types:

```{r cleanEventTypes}
# everything to upper case
data$EVTYPE <- toupper(data$EVTYPE)

# TSTM -> THUNDERSTORM
data$EVTYPE <- gsub("TSTM", "THUNDERSTORM", data$EVTYPE)

# combine types containing a certain keyword into a single event type
keys <- c("HURRICANE", "TORNADO", "THUNDERSTORM WIND", "FLASH FLOOD", "BLIZZARD", "HEAVY RAIN", "HEAVY SNOW")
for (key in keys) {
    index <- grep(key, data$EVTYPE)
    data$EVTYPE[index] <- key
}

data$EVTYPE <- factor(data$EVTYPE)
```

The number of event types is now `r length(levels(data$EVTYPE))`, with the top five:
```{r eventTypes2}
head(summary(data$EVTYPE), 5)
```

A much more thorough cleaning-up of event types in the data would be needed, but is beyond the scope of this assignment.


### Financial damage
```{r processDamage}
expName <- c("U", "K", "M", "B")
expNumber <- c(1, 1E3, 1E6, 1E9)
names(expNumber) <- expName

```
