---
title: "Storm Data"
author: "patmau"
date: "6/11/2020"
output: html_document
---

# Most Harmful Weather Events in the US with Respect to Population Health and Economical Damage

## Synopsis

The NOAA storm database collects information about major storms and weather events in the United States, including estimates of fatalities, injuries, and economical damage.
In this assignment, we look at events in the database from 1950 to November 2011, and try to determine which type of events are most harmful to population health and cause the highest economical damage.

We find that in terms of fatalities, tornadoes are the most harmful events, followed by excessive heat and flash floods.
Tornadoes also cause the highest number of injuries by far, followed by thunderstorm winds and floods.
The highest economical damage is caused by hurricanes, followed by tornadoes and storm surges.
Tornadoes should thus probably be considered the most dangerous type of weather event, as they cause large damage to population health as well as large economical damage.
They are also the third most frequent event in the database.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(gridExtra)
library(formattable)
```

## Download and process data

Download the data file if necessary and load it into R:

```{r download, cache=TRUE}
dataFile <- "data/StormData.csv.bz2"
if (!file.exists(dataFile)) {
    url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    download.file(url, dataFile, method = "curl")
    date()
}

data <- read.csv(dataFile, header = TRUE, stringsAsFactors = FALSE)
nEvents <- nrow(data)
```
The data set contains a total of `r nEvents` events.

### Event types

The [Storm Data Documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf) lists 48 permitted event types.
In the data file, however, we find no less than `r length(levels(factor(data$EVTYPE)))` event types.
The top five most occurring event types:
```{r eventTypes}
head(summary(factor(data$EVTYPE)), 5)
```
Clearly, "TSTM WIND" and "THUNDERSTORM WIND" refer to the same type of event.
We perform a coarse clean-up of event types:

```{r cleanEventTypes}
# everything to upper case
data$EVTYPE <- toupper(data$EVTYPE)

# TSTM -> THUNDERSTORM
data$EVTYPE <- gsub("TSTM", "THUNDERSTORM", data$EVTYPE)

# combine types containing a certain keyword into a single event type
keys <- c("HURRICANE", "TORNADO", "THUNDERSTORM WIND", "FLASH FLOOD", "BLIZZARD", "HEAVY RAIN", "HEAVY SNOW")
for (key in keys) {
    index <- grep(key, data$EVTYPE)
    data$EVTYPE[index] <- key
}

data$EVTYPE <- factor(data$EVTYPE)
```

The number of event types is now `r length(levels(data$EVTYPE))`, with the top five:
```{r eventTypes2}
head(summary(data$EVTYPE), 5)
```

A much more thorough cleaning-up of event types would be needed, but is beyond the scope of this assignment.


### Financial damage

Financial damage results from damage to property and damage to crop.
We combine the types of damage into a single value for total financial damage caused by a given event.
Damage is represented in the dataset as a combination of a number ("`PROPDMG`") and a magnitude expressed as an alphabetical character (`PROPDMGEXP`).
According to the documentation, allowed characters include 'K', 'M', 'B', for 10^3, 10^6, and 10^9.
We find:
```{r damageExponents}
summary(factor(data$PROPDMGEXP))
summary(factor(data$CROPDMGEXP))
```
The vast majority of entries contain either no exponent or a valid one, and only those will be considered in this analysis. We will assume that a missing exponent corresponds to a factor 10^0 = 1.

```{r processDamage}
# valid exponents
expName <- c("U", "K", "M", "B")
expValue <- c(1, 1E3, 1E6, 1E9)
names(expValue) <- expName

# replace empty exponents with "U" (Unit) 
data$PROPDMGEXP[data$PROPDMGEXP == ""] <- "U"
data$CROPDMGEXP[data$CROPDMGEXP == ""] <- "U"

# extract entries with valid exponents
validExp <- data$PROPDMGEXP %in% expName & data$CROPDMGEXP %in%expName
damageData <- data[validExp, ]

# combine property and crop damage
damageData$totalDamage <- damageData$PROPDMG * expValue[damageData$PROPDMGEXP] +
    damageData$CROPDMG * expValue[damageData$CROPDMGEXP]

nEventsWithDamage <- nrow(damageData)
```
We obtain `r nEventsWithDamage` events with a valid value for total damage, which is `r percent(nEventsWithDamage / nEvents)` of all recorded events.

The single event with the largest total damage:
```{r maxDamage}
maxEvent <- which.max(damageData$totalDamage)
damageData[maxEvent, ]
```
A total damage of more than 100 Billion USD seems excessive for this event. It is more than three times the damage attributed to storm surge in Louisiana during Hurricane Katrina (this event ranks 2nd).
For the Napa Valley flooding event, we will set the total damage to 100M USD, which seems a more reasonable value (the `REMARKS` entry for this event mentions a damage of at least 70M).
```{r napaDamage}
damageData$totalDamage[maxEvent] <- 100E6
```


## Results

### Economical consequences

We assess economical consequences by computing the financial damage caused by an event type, accumulated over the entire time period in the data set (1950 - Nov 2011). Note that inflation is not accounted for. 

```{r financialDamage}
damage <- select(damageData, EVTYPE, totalDamage) %>%
    group_by(EVTYPE) %>%
    summarise(total = sum(totalDamage),
              count = n(),
              mean = mean(totalDamage)) %>%
    arrange(desc(total))

top5 <- head(damage, 5)

ggplot(data = top5,
       aes(x = reorder(EVTYPE, -total),
           y = total * 1E-9,
           fill = log10(count))) +
    geom_col() +
    labs(title = "Total Damage 1950 - 2011 by Event Type",
         x = "",
         y = "Damage in billion USD",
         fill = "Nr. of events\n(log scale)")

```

We find that hurricanes cause the largest financial damage overall, followed by tornadoes.
The coloring according to number of events suggests that we can discern two broad categories of events:

1. Relatively rare events causing very large damage (hurricane, storm surge)
2. Events that cause small to moderate damage, but occur much more frequently (tornado, flood, hail)

A table including mean damage per event:

```{r meanDamage}
top5Table <- data.frame(top5$EVTYPE)
top5Table <- mutate(top5Table,
                    mean = digits(top5$mean * 1E-6, 2),
                    count = top5$count,
                    total = digits(top5$total * 1E-6, 0))
names(top5Table) <- c("Event", "Mean Damage [million USD]", "Nr. of events", "Total Damage [million USD]")
formattable(top5Table,
            align = c("l", "r", "r", "r")
)
```
The table above supports our assumption. Hurricanes and storm surges are relatively rare, but cause hundreds of millions of dollars of damage on average. A hail event, on the other hand, results in a mean damage of only 60k USD. However, the sheer number of events results in a total damage in the billions.

### Population health

We assess the consequences for population health by computing the total number of fatalities and injuries (separately) caused by events:

```{r healthDamage}
casualties <- select(data, EVTYPE, FATALITIES, INJURIES) %>%
    group_by(EVTYPE) %>%
    summarise(injured = sum(INJURIES),
              fatal = sum(FATALITIES),
              count = n())

top5fatal <- head(arrange(casualties, desc(fatal)), 5)
top5injured <- head(arrange(casualties, desc(injured)), 5)

#line breaks for x-axis labels
top5fatal$EVTYPE <- gsub(" ", "\n", top5fatal$EVTYPE)
top5injured$EVTYPE <- gsub(" ", "\n", top5injured$EVTYPE)

fatalplot <- ggplot(data = top5fatal,
       aes(x = reorder(EVTYPE, -fatal),
           y = fatal,
           fill = log10(count))
       ) +
    labs(title = "Total Number of Fatalities 1950 - 2011",
         x = "",
         y = "Fatalities",
         fill = "Nr. of events\n(log scale)") +
    geom_bar(position = "stack", stat = "identity") 

injuredplot <- ggplot(data = top5injured,
       aes(x = reorder(EVTYPE, -injured),
           y = injured,
           fill = log10(count))
       ) +
    labs(title = "Total Number of Injuries 1950 - 2011",
         x = "",
         y = "Injuries",
         fill = "Nr. of events\n(log scale)") +
    geom_bar(position = "stack", stat = "identity") 

grid.arrange(fatalplot, injuredplot, nrow = 2)
```

The most harmful events by far are tornadoes in both categories. In terms of fatalities, excessive heat ranks second, while the second most injuries are caused by thunderstorm wind. Notably, hurricanes cause huge economical damage, but do not appear in either top five in terms of health damage.
Thunderstorm winds are responsible for the second most injuries, due to the large number of events; similar to hail when it comes to financial damage.
In fact, thunderstorm winds and hail are the most frequent events in the data set (followed by tornadoes) and account for more than 60% of all events:

```{r frequentEvents}
events <- select(data, EVTYPE) %>%
    group_by(EVTYPE) %>%
    summarise(count = n()) %>%
    arrange(desc(count))

top3Event <- head(events, 3)
top3Event$percent <- percent(top3Event$count / nEvents)
names(top3Event) <- c("Event", "Nr. of Events", "Percentage of all Events")
formattable(top3Event,
            align = c("l", "r", "r"))
```

## System info

```{r systemInfo}
sessionInfo()
```
